---
title: "Visualizations for project"
author: "Mutasim Mahmud, 0813161"
date: "`r Sys.Date()`"
output: pdf_document
---
Necessary setup to not show any wanrning signs,
```{r setup, include=FALSE}
# Using this code to Remove warnings while running R chunks
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE,eval=TRUE)
```

First we will load the dataset,
```{r}
library(dplyr)
data <- read.csv("C:\\Local Disk (A)\\Trent University\\Data Visualization\\Project\\Final Project\\HR_Analytics.csv")
head(data,5)
```

See the attributes and their structure,
```{r}
str(data)
```

Cheking other characteristics of the data,
```{r}
summary(data)
```

Now we will do visualizations!

1.Trellis Plot
```{r}
library(ggplot2)

# Loading the dataset
data <- read.csv("C:\\Local Disk (A)\\Trent University\\Data Visualization\\Project\\Final Project\\HR_Analytics.csv")

# Converting necessary attributes to factors
data$AgeGroup <- as.factor(data$AgeGroup)
data$Attrition <- as.factor(data$Attrition)

# Trellis Plot
ggplot(data, aes(x=AgeGroup, y=JobSatisfaction, fill=Attrition)) + geom_boxplot() + facet_wrap(~ Attrition) + labs(title="Job Satisfaction by Age Group and Attrition", x="AgeGroup", y="JobSatisfaction") + theme(axis.text.x = element_text(angle=0, hjust=2))

```
In No group, Most employees have a Job Satisfaction level between 3 and 4, regardless of their age group. In yes group, Younger age groups (18-25 and 26-35) have a wider range of Job Satisfaction levels, indicating more variability in their satisfaction. Higher Job Satisfaction levels are generally associated with employees who did not leave the company. There may be a potential relationship between lower Job Satisfaction and higher attrition rates, particularly among younger employees.

2.
```{r}
library(ggplot2)

# Loading the dataset
data <- read.csv("C:\\Local Disk (A)\\Trent University\\Data Visualization\\Project\\Final Project\\HR_Analytics.csv")

# Converting necessary attributes to factors
data$Attrition <- as.factor(data$Attrition)
data$YearsSinceLastPromotion <- as.factor(data$YearsSinceLastPromotion)

# stacked bar plot
ggplot(data, aes(x=(YearsSinceLastPromotion), fill=Attrition)) +
  geom_bar(position="stack") +
  labs(title="Years Since Last Promotion by Attrition", x="Years Since Last Promotion", y="Count") +
  theme_minimal()

```
A significant number of employees who left the company (Attrition = Yes) have 0 years since their last promotion. This indicates that many employees left shortly after being promoted.
Employees who have not been promoted for a long time (e.g., more than 4 years) generally show fewer cases of attrition. This suggests that long-term employees might be more stable and less likely to leave, even without recent promotions.
There are small peaks in attrition at 1 and 2 years since the last promotion. This could indicate critical periods where employees reassess their career progression and decide to leave if they do not see further advancement.

3.
```{r}
# Load necessary libraries
library(ggplot2)
library(reshape2)
library(dplyr)

# Load the dataset
data <- read.csv("C:\\Local Disk (A)\\Trent University\\Data Visualization\\Project\\Final Project\\HR_Analytics.csv")

# Select necessary columns for correlation
data_selected <- data %>%
  select(Education, EnvironmentSatisfaction, PerformanceRating, RelationshipSatisfaction, WorkLifeBalance, Attrition)

# Convert Attrition to numeric for correlation
data_selected$Attrition <- as.numeric(data_selected$Attrition == "Yes")

# Compute the correlation matrix
cor_matrix <- cor(data_selected, use="complete.obs")

# Melt the correlation matrix into long format
melted_cor_matrix <- melt(cor_matrix)

# Plot the heatmap
ggplot(data = melted_cor_matrix, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() +
  labs(title="Correlation Heatmap of Satisfaction Metrics and Attrition", 
       x="", y="")



```

```{r}
library(corrplot)

# Load the dataset
data <- read.csv("C:\\Local Disk (A)\\Trent University\\Data Visualization\\Project\\Final Project\\HR_Analytics.csv")

data$Attrition <- as.numeric(factor(data$Attrition))
data$BusinessTravel <- as.numeric(factor(data$BusinessTravel))
data$Department <- as.numeric(data$Department)
data$EducationField <- as.numeric(data$EducationField)
data$Gender <- as.numeric(factor(data$Gender))
data$JobRole <- as.numeric(data$JobRole)
data$MaritalStatus <- as.numeric(data$MaritalStatus)
data$SalarySlab <- as.numeric(data$SalarySlab)
data$Over18 <- as.numeric(factor(data$Over18))
data$OverTime <- as.numeric(factor(data$OverTime))


correlation_matrix <- cor(core.data[c("Age","AgeGroup","Attrition
","BusinessTravel","DailyRate","Department","DistanceFromHome","Education","EducationField","EmployeeCount","EmployeeNumber","EnvironmentSatisfaction","Gender","HourlyRate","JobInvolvement","JobLevel","JobRole","JobSatisfaction","MaritalStatus","MonthlyIncome","SalarySlab","MonthlyRate","NumCompaniesWorked","Over18","OverTime","PercentSalaryHike
","PerformanceRating","RelationshipSatisfaction","StandardHours","StockOptionLevel","TotalWorkingYears","TrainingTimesLastYear","WorkLifeBalance","YearsAtCompany","YearsInCurrentRole","YearsSinceLastPromotion","YearsWithCurrManager")])
 
corrplot(correlation_matrix, method = "color", type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.6)


```

